---
title: "Diversification analysis with RPANDA"
output:
  html_document:
    fig_height: 8
    fig_width: 9
    selfcontained: yes
    toc: yes
    toc_depth: 3
  md_document:
    variant: markdown_github
  pdf_document:
    fig_height: 6
    fig_width: 7
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc: yes
    toc_depth: 3
date: '`r format(Sys.Date(), "%B %d, %Y")`'
---


# RPANDA

[RPANDA](http://cran.r-project.org/web/packages/RPANDA/index.html) is
another recently developed R package that allows us to fit different
models of rate variation through time and select the best fitting
model using maximum likelihood analysis.

RPANDA shows basically two main differences from BAMM:

1. In RPANDA, the user must inform which models are going to be
tested, whereas in BAMM the program itself will average among the
rates for each clade;

2. In RPANDA, the user must know *a priori* which are the clades that
might show a particular diversification regime, while BAMM will
estimate where rate shifts are positioned using a rjMCMC algorithm.

The package also contains some simuation functions as well as
datasets. You can find more details on [CRAN](http://cran.r-project.org/web/packages/RPANDA/RPANDA.pdf).

## Model selection

For this exercise, we will test four different scenarios with all
combinations of constant and variable speciation and extinction rates
as follows (based on
[Morlon *et al.* 2011, PNAS](http://www.pnas.org/content/108/39/16327.abstract?tab=ds))

Read in the whale tree
```{r}
library(ape)
library(RPANDA)
whales <- read.tree("whaletree.tre")
plot(whales, cex = 0.35)
nodelabels(cex = 0.5)

```


## Rate varying functions

Here we define four functions that describe how $/lambda$ and $/mu$ can change over time. We will use these functions to visualize rate change over time AND we will pass them to RPANDA to evaluate the likelihood of the whale tree under them below.

```{r}
lambda.cst <- function(x,y){y}
lambda.var <- function(x,y){y[1]*exp(y[2]*x)}
mu.cst <- function(x,y){y}
mu.var <- function(x,y){y[1]*exp(y[2]*x)}
```

The rate variations for the four possible pairwise combinations look
like this:

```{r fig.width=9, fig.height=8, echo=FALSE}
par(mfrow=c(2,2))
plot(rep(1,length(seq(0,25,0.1)))~seq(0,25,0.1), type="l", ylim=c(0,1.2), xlab="time",ylab="rate", main="BCSTDCST", lwd=2)
legend("topleft",legend=c("speciation","extinction"),col=c(1,2), lty=1, cex=0.75, lwd=2)
abline(h=0.2,col=2, lwd=2)
plot(lambda.var(seq(0,25,0.1),c(1,-0.1))~seq(0,25,0.1), type="l", xlab="time",ylab="rate", main="BVARDCST", lwd=2)
legend("topright",legend=c("speciation","extinction"),col=c(1,2), lty=1, cex=0.75, lwd=2)
abline(h=0.2,col=2, lwd=2)
plot(mu.var(seq(0,25,0.1),c(0.2,0.075))~seq(0,25,0.1), type="l",col=2, xlab="time",ylab="rate", main="BCSTDVAR", lwd=2)
legend("topleft",legend=c("speciation","extinction"),col=c(1,2), lty=1, cex=0.75, lwd=2)
abline(h=1, lwd=2)
plot(lambda.var(seq(0,25,0.1),c(1,-0.1))~seq(0,25,0.1), type="l", xlab="time",ylab="rate", main="BVARDVAR", lwd=2)
legend("topright",legend=c("speciation","extinction"),col=c(1,2), lty=1, cex=0.75, lwd=2)
lines(mu.var(seq(0,25,0.1),c(0.1,0.075))~seq(0,25,0.1), type="l",col=2, lwd=2)




```

Having decided which models one would like to test, the next step is
to extract the clades that most likely share similar diversification
regimes. As said previously, RPANDA needs the user to *a priori*
indicate these clades. In this example, we will use the four main cetacean
families plus the rest of cetaceans as a fifth clade.

```{r}

##create a pdf of the tree with the internal nodes labelled
pdf("whale_tree.pdf", width = 4, height =16)
plot(whales, cex = 0.5)
nodelabels(cex = 0.5)
dev.off()

```

```{r}
#Use these node numbers to extract clades.
balaenidae.tree  <- extract.clade(whales, 90)
balaenopteridae.tree <- extract.clade(whales,95)
delphinidae.tree <- extract.clade(whales, 140)
phocoenidae.tree <- extract.clade(whales, 135)
ziphidae.tree <- extract.clade(whales, 109)

par(mfcol = c(3, 2))
plot(balaenidae.tree, main = "Balaenidae")
plot(balaenopteridae.tree, main = "Balaenopteridae") 
plot(delphinidae.tree, main = "Delphinidae")
plot(phocoenidae.tree, main = "Phocoenidae")
plot(ziphidae.tree, main = "Ziphidae")

#We also need to capture the rest of the cetaceans that do not fall into these clades

othercetaceans.tree <- drop.tip(whales,c(balaenopteridae.tree$tip.label,delphinidae.tree$tip.label, phocoenidae.tree$tip.label, ziphidae.tree$tip.label))

plot(othercetaceans.tree, main = "Other cetaceans")


```

If you do not know the node numbers, you can also extract clades by passing a list of clade names. This block shows how we could use grep to accomplish this.
```{r, eval=FALSE}

balaenidae <- c(whales$tip.label[grep("Balaena",whales$tip.label)],whales$tip.label[grep("Eubalaena",whales$tip.label)])
balaenopteridae <- c(whales$tip.label[grep("Balaenoptera",whales$tip.label)],whales$tip.label[grep("Megaptera",whales$tip.label)])
delphinidae <- c(whales$tip.label[grep("Delphinus",whales$tip.label)],whales$tip.label[grep("Cephalorhynchus",whales$tip.label)],whales$tip.label[grep("Feresa",whales$tip.label)],whales$tip.label[grep("Globicephala",whales$tip.label)],whales$tip.label[grep("Lagenodelphis",whales$tip.label)],whales$tip.label[grep("Lagenorhynchus",whales$tip.label)],whales$tip.label[grep("Lissodelphis",whales$tip.label)],whales$tip.label[grep("Orcaella",whales$tip.label)],whales$tip.label[grep("Orcinus",whales$tip.label)],whales$tip.label[grep("Peponocephala",whales$tip.label)],whales$tip.label[grep("Pseudorca",whales$tip.label)],whales$tip.label[grep("Sotalia",whales$tip.label)],whales$tip.label[grep("Sousa",whales$tip.label)],whales$tip.label[grep("Stenella",whales$tip.label)],whales$tip.label[grep("Steno",whales$tip.label)],whales$tip.label[grep("Tursiops",whales$tip.label)],whales$tip.label[grep("Grampus",whales$tip.label)])
phocoenidae <- c(whales$tip.label[grep("Neophocaena",whales$tip.label)],whales$tip.label[grep("Phocoena",whales$tip.label)],whales$tip.label[grep("Phocoenoides",whales$tip.label)])
ziphidae <- c(whales$tip.label[grep("Berardius",whales$tip.label)],whales$tip.label[grep("Hyperoodon",whales$tip.label)],whales$tip.label[grep("Indopacetus",whales$tip.label)],whales$tip.label[grep("Mesoplodon",whales$tip.label)],whales$tip.label[grep("Tasmacetus",whales$tip.label)],whales$tip.label[grep("Ziphius",whales$tip.label)])
balaenidae.tree <- drop.tip(whales,whales$tip.label[-match(balaenidae,whales$tip.label)])
```

```{r, eval=FALSE}
balaenopteridae.tree <- drop.tip(whales,whales$tip.label[-match(balaenopteridae,whales$tip.label)])
delphinidae.tree <- drop.tip(whales,whales$tip.label[-match(delphinidae,whales$tip.label)])
phocoenidae.tree <- drop.tip(whales,whales$tip.label[-match(phocoenidae,whales$tip.label)])
ziphidae.tree <- drop.tip(whales,whales$tip.label[-match(ziphidae,whales$tip.label)])
othercetaceans.tree <- drop.tip(whales,c(balaenopteridae,delphinidae, phocoenidae, ziphidae))

```

Lets fit our four models to the tree for the phocoenidae.


```{r}
#fit a constant rate bd model, bvardcst
fit_bd(phocoenidae.tree, max(branching.times(phocoenidae.tree)), f.lamb=lambda.cst, f.mu=mu.cst, lamb_par=0.4, mu_par=0,cst.lamb=TRUE,cst.mu=TRUE,cond="crown",f=87/89,dt=1e-3)

#how does this compare to the bd fit under ape from this AM?
bd<-function(x){
    if(class(x)!="birthdeath") stop("x should be an object of class 'birthdeath'")
    b<-x$para[2]/(1-x$para[1])
    d<-b-x$para[2]
    setNames(c(b,d),c("b","d"))
}

fit.bd<-birthdeath(phocoenidae.tree)
bd(fit.bd)

# diversitree fit
# library(diversitree)
# pbModel<-make.bd(phocoenidae.tree,sampling.f=87/89)
# bdMLFit<-find.mle(bdModel,c(0.1,0.05),method = "optim",lower = 0)
# bdMLFit

##fit a model where the speciation rate varies through time and the extinction rate is constant, bvardcst
fit_bd(phocoenidae.tree, max(branching.times(phocoenidae.tree)), f.lamb=lambda.var, f.mu=mu.cst, lamb_par= c(0.4,-0.05),mu_par=0,expo.lamb=TRUE,cst.mu=TRUE,cond="crown",f=87/89,dt=1e-3)


```




This function will help automate the calculation of the likelihood for the RPANDA models.

```{r}
library(RPANDA)
fit.multi.rpanda <- function(tree,par)
    {
        bcstdcst <- fit_bd(tree, max(branching.times(tree)), f.lamb=lambda.cst, f.mu=mu.cst, lamb_par=par[[1]][1],mu_par=par[[1]][2],cst.lamb=TRUE,cst.mu=TRUE,cond="crown",f=87/89,dt=1e-3)
        bvardcst <- fit_bd(tree, max(branching.times(tree)), f.lamb=lambda.var, f.mu=mu.cst, lamb_par=par[[2]][c(1,2)],mu_par=par[[2]][3],expo.lamb=TRUE,cst.mu=TRUE,cond="crown",f=87/89,dt=1e-3)
        bcstdvar <- fit_bd(tree, max(branching.times(tree)), f.lamb=lambda.cst, f.mu=mu.var, lamb_par=par[[3]][1],mu_par=par[[3]][c(2,3)],cst.lamb=TRUE,expo.mu=TRUE,cond="crown",f=87/89,dt=1e-3)
        bvardvar <- fit_bd(tree, max(branching.times(tree)), f.lamb=lambda.var, f.mu=mu.var, lamb_par=par[[4]][c(1,2)],mu_par=par[[4]][c(3,4)],expo.lamb=TRUE,expo.mu=TRUE,cond="crown",f=87/89,dt=1e-3)
        return(list("bcstdcst"=bcstdcst,"bvardcst"=bvardcst,"bcstdvar"=bcstdvar,"bvardvar"=bvardvar))
    }
whales.par <- list(c(0.4,0),c(0.4,-0.05,0),c(0.4,0.1,0.05),c(0.4,-0.05,0.1,0.05)) #we need to supply starting parameter values for optimization to RPANDA
```

## Estimation of model parameters

In the posession of all models and clades, we can finally estimate the
parameters of all the four models to each of the five clades and
create an AICc table in order to select which one is the best model
that describes the changes in both diversification rates (speciation
and extinction) rates. (This part of the code take a while to complete)

```{r}
results <- list("balaenopteridae.res"=fit.multi.rpanda(balaenopteridae.tree,whales.par),
                "delphinidae.res" = fit.multi.rpanda(delphinidae.tree,whales.par),
                "phocoenidae.res" = fit.multi.rpanda(phocoenidae.tree,whales.par),
                "ziphidae.res" = fit.multi.rpanda(ziphidae.tree,whales.par),
                "othercetaceans.res"= fit.multi.rpanda(othercetaceans.tree,whales.par))
```

```{r}
aic.table <- matrix(nrow=4,ncol=5,NA)
for(i in 1:5)
    {
        for(j in 1:4)
            {
                aic.table[j,i] <- results[[i]][[j]]$aicc
            }
    }
colnames(aic.table) <- c("Balaenopteridae","Delphinidae","Phocoenidae","Ziphidae","Other Cetaceans")
rownames(aic.table) <- c("bcstdcst","bvardcst","bcstdvar","bvardvar")
aic.table
par.table <- data.frame("Balaenopteridae"=c(results[[1]]$bcstdcst$lamb_par[1:2],results[[1]]$bcstdcst$mu_par[1:2]),"Delphinidae"=c(results[[2]]$bvardcst$lamb_par[1:2],results[[2]]$bvardcst$mu_par[1:2]),"Phocoenidae"=c(results[[3]]$bcstdcst$lamb_par[1:2],results[[3]]$bcstdcst$mu_par[1:2]),"Ziphidae"=c(results[[4]]$bcstdcst$lamb_par[1:2],results[[4]]$bcstdcst$mu_par[1:2]),"Other Cetaceans"=c(results[[5]]$bcstdvar$lamb_par[1:2],results[[5]]$bcstdvar$mu_par[1:2]))
par.table
```

## Plotting diversity through time

After selecting which model best fits the trees, we can estimate the
diversity trajectory through time for each of the five clades.

```{r}
# Function to calculate species richness in a given point in time
div.times <- c(max(branching.times(balaenopteridae.tree)),max(branching.times(delphinidae.tree)),max(branching.times(phocoenidae.tree)),max(branching.times(ziphidae.tree)),max(branching.times(othercetaceans.tree)))

# Function modified from plot_dtt from RPANDA package
plotdtt <- function (fit.bd, tot_time, N0, col=1, add=FALSE, div.time, xlim, ylim)
{
    if (!inherits(fit.bd, "fit.bd"))
        stop("object \"fit.bd\" is not of class \"fit.bd\"")
    t <- seq(tot_time-div.time, tot_time, 0.01)
    if ("f.mu" %in% attributes(fit.bd)$names) {
        r <- function(t) {
            -fit.bd$f.lamb(t) + fit.bd$f.mu(t)
        }
        R <- function(s) {
            RPANDA:::.Integrate(Vectorize(r), 0, s)
        }
        N <- N0 * exp(Vectorize(R)(t))
                                        #dev.new()
        if(add==FALSE)
            {
        plot(-t, N, type = "l", xlab = "time", ylab = "Number of species",
             main = "Diversity Through Time", col=col, xlim=xlim, ylim=ylim)
    }
        else
            {
                lines(-t, N, type = "l", xlab = "time", ylab = "Number of species",
                     main = "Diversity Through Time", col=col, xlim=xlim, ylim=ylim)
            }
    }
    else {
        r <- function(t) {
            -fit.bd$f.lamb(t)
        }
        R <- function(s) {
            RPANDA:::.Integrate(Vectorize(r), 0, s)
        }
        N <- N0 * exp(Vectorize(R)(t))
                                        #dev.new()
        if(add==FALSE)
            {
        plot(-t, N, type = "l", xlab = "time", ylab = "Number of species",
             main = "Diversity Through Time",col=col, xlim=xlim, ylim=ylim)
    }
        else
            {
                lines(-t, N, type = "l", xlab = "time", ylab = "Number of species",
                     main = "Diversity Through Time",col=col, xlim=xlim, ylim=ylim)
            }
    }
}

plotdtt(results$balaenopteridae$bcstdcst,div.times[1],N0=Ntip(balaenopteridae.tree),xlim=c(-max(div.times),0),ylim=c(0,150),div.time=div.times[1])
plotdtt(results$delphinidae$bvardcst,div.times[2],N0=Ntip(delphinidae.tree),col=6,add=TRUE,xlim=c(-max(div.times),0),ylim=c(0,150),div.time=div.times[2])
plotdtt(results$phocoenidae$bcstdcst,div.times[3],N0=Ntip(phocoenidae.tree),col="goldenrod",add=TRUE,xlim=c(-max(div.times),0),ylim=c(0,150),div.time=div.times[3])
plotdtt(results$ziphidae$bcstdcst,div.times[4],N0=Ntip(ziphidae.tree),col=4,add=TRUE,xlim=c(-max(div.times),0),ylim=c(0,150),div.time=div.times[4])
plotdtt(results$othercetaceans$bcstdvar,div.times[5],N0=Ntip(othercetaceans.tree),col="darkred",add=TRUE,xlim=c(-max(div.times),0),ylim=c(0,150),div.time=div.times[5])
legend("topleft",legend=c("Balaenopteridae","Delphinidae","Phocoenidae","Ziphidae","Other Cetaceans"),text.col=c(1,6,"goldenrod",4,"darkred"))
```

# Challenge

1. Fit models using RPANDA for the **entire** whale phylogeny, and
   plot its diversity through time using parts of the previous code
   chunks.


